---
title: "Gene Quantification"
subtitle: "Step 3: Generating gene count matrices"
sidebar: bulk-rnaseq
---
## Overview

Gene quantification converts aligned reads into count matrices that can be used for differential expression analysis.

## featureCounts

featureCounts from the Subread package is fast and accurate for counting reads mapped to genomic features.

```{bash}
#| eval: false
#| code-fold: false

# Count reads with featureCounts
featureCounts \
    -T 8 \
    -p --countReadPairs \
    -t exon \
    -g gene_id \
    -a Homo_sapiens.GRCh38.110.gtf \
    -o gene_counts.txt \
    aligned/*.bam
```

### Parameters Explained

| Parameter | Description |
|-----------|-------------|
| `-p --countReadPairs` | Count fragments instead of reads (PE data) |
| `-t exon` | Feature type to count |
| `-g gene_id` | Attribute for grouping features |
| `-s 2` | Strand-specific (use 1 or 2 based on library) |

## Preparing Count Matrix in R

```{r}
#| eval: false
#| code-fold: false

library(tidyverse)

# Read featureCounts output
counts_raw <- read_tsv("gene_counts.txt", comment = "#")

# Clean up column names and extract counts
counts <- counts_raw |>
  select(Geneid, ends_with(".bam")) |>
  rename_with(~ str_remove(.x, "_Aligned.sortedByCoord.out.bam"), ends_with(".bam")) |>
  column_to_rownames("Geneid")

# View dimensions
dim(counts)
```

## Salmon (Alignment-Free Quantification)

Salmon provides fast, accurate transcript-level quantification without alignment.

### Build Salmon Index

```{bash}
#| eval: false

# Download transcriptome
wget https://ftp.ensembl.org/pub/release-110/fasta/homo_sapiens/cdna/Homo_sapiens.GRCh38.cdna.all.fa.gz

# Build decoy-aware index
grep "^>" Homo_sapiens.GRCh38.dna.primary_assembly.fa | cut -d " " -f 1 > decoys.txt
sed -i.bak -e 's/>//g' decoys.txt

cat Homo_sapiens.GRCh38.cdna.all.fa.gz Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz > gentrome.fa.gz

salmon index -t gentrome.fa.gz -d decoys.txt -p 12 -i salmon_index
```

### Run Salmon Quantification

```{bash}
#| eval: false
#| code-fold: false

#!/bin/bash
# Salmon quantification

for R1 in trimmed_data/*_R1_trimmed.fastq.gz; do
    SAMPLE=$(basename ${R1} _R1_trimmed.fastq.gz)
    R2=trimmed_data/${SAMPLE}_R2_trimmed.fastq.gz
    
    salmon quant \
        -i salmon_index \
        -l A \
        -1 ${R1} \
        -2 ${R2} \
        -p 8 \
        --validateMappings \
        --gcBias \
        --seqBias \
        -o salmon_quant/${SAMPLE}
done
```

## Import Salmon Results with tximport

```{r}
#| eval: false
#| code-fold: false

library(tximport)
library(tidyverse)

# Create tx2gene mapping
library(biomaRt)
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
tx2gene <- getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = mart
)

# List Salmon quantification files
files <- list.files("salmon_quant", pattern = "quant.sf", 
                    recursive = TRUE, full.names = TRUE)
names(files) <- basename(dirname(files))

# Import with tximport
txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

# Gene-level counts for DESeq2
counts <- round(txi$counts)
```

## Quality Check: Count Distribution

```{r}
#| eval: false

library(ggplot2)

# Plot library sizes
lib_sizes <- colSums(counts)

data.frame(
  sample = names(lib_sizes),
  counts = lib_sizes
) |>
  ggplot(aes(x = reorder(sample, counts), y = counts / 1e6)) +
  geom_col(fill = "#3498db") +
  coord_flip() +
  labs(
    title = "Library Sizes",
    x = "Sample",
    y = "Total Counts (millions)"
  ) +
  theme_minimal()
```

## Filtering Low-Expression Genes

```{r}
#| eval: false
#| code-fold: false

# Filter genes with low counts
# Keep genes with at least 10 counts in at least 3 samples
keep <- rowSums(counts >= 10) >= 3
counts_filtered <- counts[keep, ]

# Report filtering
message(sprintf("Genes before filtering: %d", nrow(counts)))
message(sprintf("Genes after filtering: %d", nrow(counts_filtered)))
message(sprintf("Genes removed: %d (%.1f%%)", 
                nrow(counts) - nrow(counts_filtered),
                100 * (1 - nrow(counts_filtered) / nrow(counts))))
```

## Next Steps

With the count matrix prepared, proceed to [Differential Expression Analysis](04-differential-expression.qmd).
